---
title: Example Core Session
description: Just an example
date: last-modified
image: ../img/under_construction.svg
categories:
  - Core
  - Example
  - For fun
---

```{r setup}
#| include: false
#| eval: true

solutions <- TRUE
```

## Text Formatting

1. *Italic* and **Bold** will turn out like this

2. Block quotes will look like this:

> This is a `blockquote` made using `>` 

3. Tooltips can be done using a classic html style:
<span class='hoverinfo' data-toggle='tooltip' title='test'>example</span>

4. If relevant, you might also want to use a footnote^[example footnote] instead of a tooltip.

5. Content can also go in the margin -->

::: {.column-margin}

Hi! I'm a side note!

:::

6. Finally, you can fold entire sections:
<details>
  Surprise ! 
</details>


## Code
Inline coding will turn out `like this`

Code blocks will appear like this:
```{r}
print('hello world')
```

**Warning:** For these tutorials, code blocks are **not evaluated** by default. If you want to evaluate them, you must indicate it specifically.
```{r}
#| eval: TRUE

print('hello back!')
```

Additionally, **solution** code blocks are not included by default, including them is done by setting the `solutions` parameter (in the YAML) to `TRUE` when rendering.
```{r include=solutions}
#| eval: TRUE

print('hello back!')
```

Note, this website has **code linking** activated, so you can click on functions to get their documentation.


## Callouts
Things are a bit quick and dirty at the moment. Ultimately I will make nice new classes in a full **repicentre** theme that can be used for exercised etc. For the moment we will just frankenstine some existing callouts.

::: {.callout-note}
**This is a callout** using `{.callout-note}`
:::

::: {.callout .callout-look}
**Look.** This callout can be used for "observation activities" (ie: look at this thing in the console). Eventually it will be more beautiful, just make sure you use this class. 

**WARNING.** For the moment this class looks just like the "code activities" class, please make sure you use the right one.
:::

::: {.callout .callout-code}
**Code.** This callout can be used for "code activities". Eventually it will be more beautiful, just make sure you use this class. 

**WARNING.** For the moment this class looks just like the "observation activities" class, please make sure you use the right one.
:::




## Tabsets
::: {.panel-tabset .nav-pills}

## One
Content that will show under the first tab

## Two
Content that will show under the second tab

## Three
Content that will show under the third tab
:::


## Images
You can insert images by referring to their relative path using markdown syntax or HTML. Note that the markdown syntax does not allow you to modify image size. In either case, make sure to add alt text for accessibility.

Markdown style syntax:

![example image alt text](../img/under_construction.svg)

HTML style syntax (with specification of desired size):

<img src='../img/under_construction.svg' alt='example image alt (html)' width='200'/>

