---
title: Session Title
description: Short description (appears on cards in pathway / explore pages)
date: last-modified
image: ../img/under_construction.svg
categories:
  - Core/Satellite
  - Category 1
  - Category 2
---

```{r setup}
#| include: false
#| eval: true
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```


Objectives
----------------------------------------------------------------------------------------------------
- Objective 1
- Objective 2


Setup
----------------------------------------------------------------------------------------------------
::: {.setup}
Description of something participants need to setup, primarily used at the beginning of a section but can also be used for tasks like setting up an `Rproject` file, folder structure, etc.
:::


Main Section
----------------------------------------------------------------------------------------------------
Your main section can (and probably should) be boken down into subsections.


### Subsection 1
```{r}
print('some example code')
```

Here are the validated callouts that you can use, **please do not use any callouts not defined here**. If you think there is an important callout that is missing, please [submit an issue](https://github.com/epicentre-msf/repicentre/issues).

### Subsection 2
::: {.callout-note}
Note about something of interest, like helpful hint or a relevant warning.
:::

::: {.callout-tip}
Comment about a genral tip / trick or best practice.
:::

::: {.callout-important}
Warning / comment on something really important.
:::

Have some other information that doesn't quite warrant the main text or a callout? Consider using a [tooltip]{.hovertip bs-toggle='tooltip' bs-title='Example tooltip'}.


::: {.look}
Description of something participants should observe, investigate, etc.
:::

...lorem ispum again !

::: {.write}
Description of a coding exercise that participants are expected to complete.
:::

