---
title: Session Title
description: Short description (appears on cards in pathway / explore pages)
date: last-modified
image: ../img/under_construction.svg
categories:
  - Core
---

```{r setup}
#| include: false
#| eval: true
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```


Objectives
----------------------------------------------------------------------------------------------------
- Objective 1
- Objective 2


Setup
----------------------------------------------------------------------------------------------------
::: {.setup}
Description of something participants need to setup, primarily used at the beginning of a section but can also be used for tasks like setting up an `Rproject` file, folder structure, etc.
:::


Main Section 1
----------------------------------------------------------------------------------------------------
Your main section(s) can (and probably should) be boken down into subsections.


### Subsection 1

### Subsection 2

Done!
----------------------------------------------------------------------------------------------------
This last header let's students know that they are done with the main material for the day. It should also include a link to the solutions (hosted on github). 


Going Further
----------------------------------------------------------------------------------------------------
After your main content is done you should have a section called called "Going Further" for students who finish the main content early. It should include:
1. A mention of one or two satellite sessions that would be relevant extensions of the current material
2. A section with "extra exercise questions" (these don't need to use the "action blocks" (see below) and can just be a number list as shown below.

### Extra Exercises
1. Do this.
2. Then do that.

<hr>
<hr>
<hr>

Markdown Reminders
----------------------------------------------------------------------------------------------------
The rest of this document is a reminder on qmd syntax and a basic style guide. Enjoy.

### Text Formatting

1. *Italic* and **Bold** will turn out like this

2. Block quotes will look like this:

> This is a `blockquote` made using `>` 

3. [Tooltips]{.hovertip bs-toggle='tooltip' bs-title='Example tooltip'} can be done using spans (please do not use asides or footnotes) 


### Code
Inline coding will turn out `like this`

Code blocks will appear like this:
```{r}
# comment
print('hello world')
```

**Warning:** For these tutorials, code blocks are **not evaluated** by default. If you want to evaluate them, you must indicate it specifically.
```{r}
#| eval: TRUE

# comment
print('hello back!')

test <- function(x) {
  if (x > 1) {
    return(x)
  } else {
    print('nothing to see here')
  }
}
```

**Note.** We are no longer using solution blocks, instead a single code file will be available at the end of each session contiaining code that runs through all the exercises. 


### Callouts
**IMPORTANT: please do _not_ use callouts not explicitly defined here; they have not been included in the css and therefore will not render well in the final document.** 


::: {.callout-note}
This is a callout using `{.callout-note}`
:::

::: {.callout-tip}
Comment about a genral tip / trick or best practice.
:::

::: {.callout-important}
Warning / comment on something really important.
:::


### Action Boxes
These are used for things participants are expected to *actually do*, ie: exercises. They are split into three categories.

::: {.setup}
Description of something participants need to setup, primarily used at the beginning of a section but can also be used for tasks like setting up an `Rproject` file, folder structure, etc.
:::

::: {.look}
Description of something participants should observe, investigate, etc.
:::


::: {.write}
Description of a coding exercise that participants are expected to complete.
:::



### Tabsets
::: {.panel-tabset .nav-pills}

## One
Content that will show under the first tab

## Two
Content that will show under the second tab

## Three
Content that will show under the third tab
:::


### Images
You can insert images by referring to their relative path using markdown syntax or HTML. Note that the markdown syntax does not allow you to modify image size. In either case, make sure to add alt text for accessibility.

Markdown style syntax:

![example image alt text](../img/under_construction.svg)

HTML style syntax (with specification of desired size):

<img src='../img/under_construction.svg' alt='example image alt (html)' width='200'/>


### Linking to Other Pages
Easy, use relative paths within a standard href, ie: [link to home page](../index.html).

