---
title: Session Title
description: Short description (appears on cards in pathway / explore pages)
date: last-modified
image: ../img/under_construction.svg
params: 
  solutions: FALSE
  sub_title: "**------ With Solutions ------**"
categories:
  - Core/Satellite
  - Category 1
  - Category 2
---

```{r setup}
#| include: false
#| eval: true
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)

solutions <- params$solutions

# library(...)         # what does this library do?
```


Objectives
----------------------------------------------------------------------------------------------------
- Objective 1
- Objective 2


Setup
----------------------------------------------------------------------------------------------------
::: {.setup}
Description of something participants need to setup, primarily used at the beginning of a section but can also be used for tasks like setting up an `Rproject` file, folder structure, etc.
:::


Main Section
----------------------------------------------------------------------------------------------------
Lorem ipsum...

```{r}
print('some example code')
```

Here are the validated callouts that you can use, **please do not use any callouts not defined here**. If you think there is an important callout that is missing, please [submit an issue](https://github.com/epicentre-msf/repicentre/issues).

::: {.callout-note}
Note about something of interest, like helpful hint or a relevant warning.
:::

::: {.callout-caution}
Comment about best practice.
:::

::: {.look}
Description of something participants should observe, investiage, etc.
:::

...lorem ispum again !

::: {.write}
Description of a coding exercise that participants are expected to complete.
:::

**Don't forget to include a solution**
```{r include=solutions}
#| eval: TRUE
# Solution

print('hello back!')
```




