---
title: Example Satellite Session
description: Just an example
date: last-modified
author:
  - name: Author 1
    email: email.author.1@epicentre.msf.org
  - name: Author 2
    email: email.author.2@epicentre.msf.org
categories:
  - Satellite
  - Example
---

```{r setup}
#| include: false

# library(fs)         # work with path
# library(here)       # create relative paths
# library(rio)        # import files
# library(janitor)    # data cleaning
# library(lubridate)  # date handling
# library(tidyverse)  # data science 

```

## Text Formatting

1. *Italic* and **Bold** will turn out like this

2. Block quotes will look like this:

> This is a `blockquote` made using `>` 

3. Tooltips are trickier to use than in a normal... for the moment we can perhaps use footnotes^[example footnote "tooltip"].

4. Content can also go in the margin, this is another option to get around the tooltip issue -->

::: {.column-margin}

Hi! I'm a side note!

:::

## Code
Inline coding will turn out `like this`

Code blocks will appear like this:
```{r}
print('hello world')
```

Note, this website has **code linking** activated, so you can click on functions to get their documentation.


## Callouts

::: {.callout-note}
**This is a Note** using `{.callout-note}`
:::

::: {.callout-caution}
**This is a warning** using `{.callout-caution}`. They can be collapsed using `collapse="true"` 
:::

::: {.callout-important}
**This is an important note** using `{.callout-important}`
:::



## Tabsets
::: {.panel-tabset .nav-pills}

## One
Content that will show under the first tab

## Two
Content that will show under the second tab

## Three
Content that will show under the third tab
:::


## Images
You can insert images by referring to their relative path using markdown syntax or HTML. Note that the markdown syntax does not allow you to modify image size. In either case, make sure to add alt text for accessibility.

Markdown style syntax:

![example image alt text](../img/under_construction.svg)

HTML style syntax (with specification of desired size):

<img src='../img/under_construction.svg' alt='example image alt (html)' width='200'/>

