---
title: Importing CSV Files
description: Importing csv files into R.
date: last-modified
image: ../img/under_construction.svg
categories:
  - Satellite
  - Data Import
---


```{r setup}
#| include: false
#| eval: true
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```


## Objectives


## Setup

::: {.setup}
Add text here about any packages or datasets they might need.
:::

## Main Section 1

We can import a `.csv` file in exactly the same way as an `.xlsx` file, we need only change the name of our file:

```{r}
import(file = here("data", "raw", "exemple_linelist.csv"))
```

`.csv` files sometimes use different <span tooltip-position="top" tooltip="A delimiter is the character used in a csv file to separate values. Many countries use a comma (,), but some use a semicolon (;).">delimiters</span> and <span tooltip-position="top" tooltip="Some countries use a decimal point (.) to indicate a number with a decimal, others will use a comma (,). Typically, countries using a comma for decimals will use a semicolon as a delimiter.">decimal characters</span> depending on where the file comes from. You can indicate to `import()` which characters to use for these using the `sep` and `dec` arguments respectively:

```{r}
import(file = here("data", "raw",, "exemple_linelist.csv"),
       sep = ";",
       dec = ",")
```

::: {.write}
Import the file `msf_linelist_moissala_2023-09-24.csv` from the raw data folder and store it in an object called `df_linelist_csv`.
:::


