---
title: Surveillance
description: Companion satellite to the surveillance Fetch-R module
# image: ../img/core/06_epicurves/logo.svg
categories:
  - Satellite
  - Surveillance
---

```{r setup}
#| include: false
#| eval: true
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)

library(here)        # For paths
library(rio)         # Import and export data
library(tidyverse)   # Data manipulation
library(zoo)  

# Surveillance data
data_surv_raw <- import(here("data", "surveillance_module", "raw", 
                             "data_ids_week20_2022_en.xlsx"))

# Laboratory data
data_lab_raw <- import(
  here("data", "surveillance_module", "raw", "data_labo_week20_2022_en.xlsx"), 
  skip = 7  # Skip the first seven lines
) 


data_surv <- data_surv_raw |> 
  mutate(
    
    # Format strings to lower case
    country     = tolower(country),
    province    = tolower(province),
    health_zone = tolower(health_zone),
    disease     = tolower(disease),
    
    
    # Remove excess spaces with the str_squish() function from 
    # the stringr package, very usefull!
    province    = str_squish(province),
    health_zone = str_squish(health_zone),
    
    
    # Remove spaces or "-" with the str_replace() function from 
    # the stringr package
    province = str_replace(province, pattern = "-", replacement = "_"), 
    province = str_replace(province, pattern = " ", replacement = "_"),
    health_zone = str_replace(health_zone,   pattern = "-", replacement = "_"), 
    health_zone = str_replace(health_zone,   pattern = " ", replacement = "_")
  )

data_lab <- data_lab_raw |> 
  mutate(
    
    # Clean strings
    health_zone = tolower(health_zone),    # Format strings to lower case
    health_zone = str_squish(health_zone), # Remove excess spaces
    health_zone = str_replace(health_zone, "-", "_"),  # Replace - by _
    health_zone = str_replace(health_zone, " ", "_"),  # Replace space by _
    
    # Recode igm modalities
    igm_measles = case_when(
      igm_measles == 'pos' ~ 'positive', 
      igm_measles == 'neg' ~ 'negative', 
      .default = igm_measles),
    
    igm_rubella = case_when(
      igm_rubella == 'pos' ~ 'positive', 
      igm_rubella == 'neg' ~ 'negative', 
      .default = igm_rubella)
  )

data_surv_weeks <- data_surv |> 
  select(province, health_zone, week, totalcases, totaldeaths) |>
  complete(
    nesting(province, health_zone),
    week = seq(min(data_surv$week, na.rm = TRUE), 
                 max(data_surv$week, na.rm = TRUE)),
    fill = list(totalcases = 0, 
                totaldeaths = 0)
  ) 


# Analyse
data_alert <- data_surv_weeks |>
  
  # Select the 4 health zones
  filter(health_zone %in% c("dilolo", "kowe" ,"kampemba", "lwamba")) |>
  
  # Order by provinceince, health_zone and week
  arrange(province, health_zone, week) %>%
  
  # Binary indicator for 20 cases (weekly indicator by health zone)
  mutate(
    cases20 = case_when(
      totalcases >= 20 ~ 1, 
      .default = 0)) |>
  
  
  # Cumulative indictors, need to be calculated by health zone
  mutate(
    
    # Group by provinceince and health zone
    .by = c(province, health_zone),
    
    # Cumulative cases over 3 week window (zoo::rollapply)
    cumcas = rollapply(totalcases, 
                       width = 3,        # Window width
                       sum,              # function to apply
                       na.rm = TRUE,     # Arguments to pass to the function (here, sum)
                       align = "right",  
                       partial = TRUE),
    
    # Binary indicator for 35 cumulative cases
    cumcases35 = case_when(cumcas >= 35 ~ 1, 
                           .default = 0),
    
    
    # Combined alert indicator
    # The operator | is a logical OR. Here we test:
    # is cases20 equal to 1 OR is cumcases35 equal to 1
    alert = case_when(
      (cases20 == 1 | cumcases35 == 1) ~ 1, 
      .default = 0)
  )

hz_alert <- data_alert |>
  filter(week == 20)  |>
  filter(alert == 1) |>
  pull(health_zone)
```

## Objectives

- Reuse skills aquired in the FETCH-R modules (import, clean and visualize data)
- More specifically, analyze **alert data** to help decide which alerts to prioritize for further field investigation.

## Introduction

This satellite is a companion to the case study _Measles emergency response in the Katanga region (DRC)_ from the FETCH Surveillance module and may thus not make sense as a standalone document.

From an R point of view, this tutorial builds on skills acquired throughout the FETCH-R modules, introduces a couple of useful generalist functions, and some more specialized ones. 

::: {.callout-tip}
Do not hesitate to refer to past sessions and your own scripts to remind yourself of some functions!
:::


## Setup (Question 2)

Since this is part of a specific module, you will create a new RStudio project. We refer you [to the main session](../sessions_core/02_import_data.qmd) for help creating a project and importing your data.

### Setup a new project

:::  {.setup}
1. Create a folder `surveillance_case_study` associated with the FETCH Surveillance module. Add the following subfolders in it:

- 📁 data
  - 📁 clean
  - 📁 raw
- 📁 R
- 📁 outputs

2. Create an [RStudio project](../sessions_core/02_import_data.qmd#rstudio-projects) at the root of the `surveillance_case_study` folder.

3. Download the raw data.

```{r}
#| echo: false
#| eval: true

downloadthis::download_link(
  link = 'https://github.com/epicentre-msf/repicentre/raw/main/data/clean/moissala_linelist_clean_EN.rds',
  button_label = "Download raw data",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

<br>
4. Unzip the archive and save the two Excel files it contains in the subfolder `data/raw`. 
<br>
5. Create a new script called `import_clean.R` and save it in the `R` subdirectory. Add a section to load the following packages: `{here}`, `{rio}`, and `{tidyverse}`.
:::


### Import data in R

**Reminder from the case study**: you requested access to the routine surveillance data and the laboratory data to the DRC MoH. The MoH agreed to share it with you on a weekly basis. The first dataset you received is of week 20 in 2022 (the data we are working on are simulated).

:::  {.look}
If you have not done it already, open the raw data files in Excel (or another equivalent application) to inspect them.
:::

The surveillance dataset is pretty straightforward to import. The lab dataset is slightly trickier: the data headers do not start at line one. Fear not, the `skip` argument from the `import()` function is made for this situation:

```{r}
# DO NOT RUN (PSEUDO-CODE)
import(
  here("data", "raw", "example_file.xlsx"), 
  skip = 3  # Skip the first three lines and start importing from line four.
) 
```

:::  {.write}
1. Add a section to your script dedicated to data importation. 

2. Import the surveillance data and store it into a `data_surv_raw` data frame. Then, import the lab data and save it in a `data_lab_raw` data frame.

3. Verify that the import went well for both data frames (Viewer, check the dimensions or start and tail of data frames).
:::



## Cleaning (Question 2 and 3)

### Surveillance data (Q2)

Now that the data is correctly imported, we are going to perform some more checks, as usual, before a bit of cleaning.

#### Quick checks

During the case study you won't have time to **inspect** and **clean** *all* columns in the imparted time, so for now we will focus on key columns: `health_zone`, `week`, `totalcases` and `totaldeaths`.

::: {.callout-note}
If you work on this tutorial in your own time, inspect the quality of the other columns and  cross-check information of several columns. We refer you to the discussion of the case study for more checks to perform.
:::

::: {.write}
Add a section for the exploration and cleaning of the surveillance data into your script. 
<br>
Now, explore the surveillance data frame and answer the following questions:

-   What are the column names?
-   How many provinceinces are in the dataset? Is this coherent with what you expect?
-   How many health zones are in the dataset?
-   What is the range of weeks?
-   What is the min of `totalcaseses`?
-   What is the max of the `totaldeaths`?
-   Do you notice missing data for these columns? Are the strings of text clean?
:::


#### Clean strings

Now that we have a better idea of what is the state of the data, let's start cleaning. We are going to write a *cleaning pipeline* like we did in the main modules (check out your code for the end of the [cleaning modules](../sessions_core/04_data_verbs_conditional.qmd) to see an example final pipeline).


:::  {.callout-tip}
To facilitate debugging your pipeline, add commands one by one, checking each new command before adding a new one.
:::

We are going to perform a couple of actions on the columns containing text to remove potential problems:

-   transform them to lower casse
-   remove potential extra spaces
-   replace `-` and spaces by `_`.

Because you may not have the time to do all of text colums, work on the `health_zone` or the `province` column for the following instructions.

:::  {.write}
Start a cleaning pipeline with a `mutate()` that turns the chosen column to lower casse.
:::

Now, we are going to introduce two handy functions for more text cleaning. The first one is the `str_squish()` function from the `{stringr}` package ([help page here](https://stringr.tidyverse.org/reference/str_trim.html)), that removes spaces at the start or end of the strings, and replace multiple spaces in the middle of a string by a single space.

```{r}
#| eval: true

examples <- c(" Trailing spaces     ",
              "Multiple     spaces",
              " Everything     here ")

str_squish(examples)
```

The other function, `str_replace` (also from the [`{stringr}` package](https://stringr.tidyverse.org/reference/str_replace.html?q=str_replace#null)) does what you expect from its name: replace something in a string by something else. It has a `pattern` argument that take the bit of text to be replaced, and a `replacement` arguments that takes the bit of text to use as replacement:

```{r}
#| eval: true

str_replace(
  "HAUT-KATANGA",    # A string of text (or a column, if used in a mutate)
  pattern = "-",     # The bit to replace
  replacement = "_"  # The replacement
)
```

:::  {.write}
Add steps to your mutate to:

-   Remove all unwanted spaces from your chosen column
-   Change the `-` and to `_` in the column (in two steps)

The head of these columns should now be:

```{r}
#| eval: true
#| echo: false

data_surv |> select(country, province, health_zone, maladie) |> head()
```

Store the result data frame in a `data_surv` object.
:::

#### Save the clean data

:::  {.write}
Use the `{rio}` package to export `data_surv` to a `.rds` file called `IDS_clean` in the `data/clean` subfolder of your project.
:::

### Laboratory data (Q2)

We are going to follow the same steps as before for the lab data, and focus for now on the columns `health_zone`, `igm_rougeole` and `igm_rubeole`.

#### Quick checks

:::  {.write}
Perform data checks on the colums names and dimensions. What are the categories for `igm_rougeole` and `igm_rubeole`? What do you need to do to clean these columns?
:::

#### Clean and recode strings

:::  {.write}
1. Start a new cleaning pipeline to clean the lab data. As before, for one of the text column, change it to lower casse, remove the extra spaces and replace the or `-` by `_`.

2. Recode at least one of  `igm_rougeole` or `igm_rubeole` columns so that the categories are `negatif`, `positif` and `indetermine`.

3. Store the cleaner version in a `data_lab` data frame

The head of the cleaned columns should now be:

```{r}
#| eval: true
#| echo: false
data_lab |> select(health_zone, igm_rougeole, igm_rubeole) |> head(10)
```
:::



::: {.callout-tip collapse=true}
You can use the `case_when()` function to recode the IGM columns.
:::


#### Save the clean data

:::  {.write}
Export the `data_lab` data frame to a `.rds` file called `lab_clean` in the `data/clean` subfolder of your project.
:::


### Going further

This is the end of the steps for question 2! If you finished in advance and there is still time, reuse the functions we just saw to clean the other text columns in both datasets and recode both IGM column in the lab dataset.

If you still have time, perform more checks on the data:

- Display the health zone for which the numbers by age group add up to a different number than the total (if any)
- Are there any health zone for which the number of deaths is higher than the total number of cases?
- Are there duplicated lines (fully duplicated, or several values for health zone and week)?
- Are there unrealistic case numbers?


### Complete surveillance dataset (Q3)

During the case study and the data checks, you realized that some weeks are missing from the surveillance dataset. You discussed the possible reasons for it, and the associated problems. Here we are going provinceide code to create a dataset that contains all weeks (assuming that missing weeks had zero cases and deaths).

We will use the function [`complete()` from the `{tidyr}` package](https://tidyr.tidyverse.org/reference/complete.html) to add the missing lines and fill the columns containing numbers (`totalcases` and `totaldeaths`) with zeros.

Look at the simplified example below: the Kikula health zone has no row for week 2:

```{r}
#| eval: true

# Create simplified data frame for the example, with three weeks
example_df = data.frame(
  province    = rep("haut_katanga", 5),
  health_zone = c("likasi", "likasi", "likasi", "kikula", "kikula"),
  week        = c(1, 2, 3, 1, 3),
  totalcases  = c(2, 1, 3, 1, 2))

example_df
```

We use the following code to cross provinceince and health zone and make sure that all their combinations have all the possible week values. Since the weeks range from one to three in that toy example, we pass a vector with weeks ranging from one to three

```{r}
#| eval: true

# Complete the missing week in kikula
example_df |> 
  complete(
    nesting(province, health_zone),
    week = seq(1, 3),  # vector from 1 to 3
    fill = list(totalcases = 0)
  ) 
```

Now both health zones within provinceinces have values for all three weeks.

It would be good to automatically pick the week series, since the data frame is going to change every week. To do that, we can remplace hardcoded values by the smallest and largest week number in the `week` column to get the range of weeks in the dataset:


```{r}
#| eval: true

# Complete the missing week in kikula
example_df |> 
  complete(
    nesting(province, health_zone),
    week = seq(min(week, na.rm = TRUE),   # vector ranging from smallest to largest week numbers in dataset
                 max(week, na.rm = TRUE)),
    fill = list(totalcases = 0)
  ) 
```


::: {.write}
1. Start a new pipeline that takes the `data_surv` data frame and keeps only the columns `province`, `health_zone`, `week` and `total cas`. 

2. Add a new step to your pipeline and paste the following code to complete the data frame:

```{r}
complete(
  nesting(province, health_zone),
  week = seq(min(week, na.rm = TRUE), 
               max(week, na.rm = TRUE)),
  fill = list(totalcases   = 0, 
              totaldeaths = 0 # also add zero to the totaldeaths column
  )
) 
```

3. Store the result of the pipeline in a data frame called `data_surv_weeks`. The head of that data frame looks like:

```{r}
#| eval: true
#| echo: false
data_surv_weeks |> head(10)
```

4. When you are done, export that data frame to a `.rds` file called `IDS_data_weeks_clean` in the `data/clean` subfolder of your project.
:::



## Defining alerts (Question 4)

### Preparing the dataset (Q4)

We are going to carry on preparing the datasets for the analyses.


#### Subset health zone

To simplify the work, we are going to focus on four health zones: Dilolo, Kampemba, Kowe, and Lwamba.

::: {.write}
Start a new pipeline from `data_surv_weeks`. Its first step it to only retain data for the the Dilolo, Kampemba, Kowe, and Lwamba health zones. 
:::


#### Weekly indicator

The first indicator we want to caclulate is whether *a health zone has 20 or more suspected cases in one week?*. This indicator is binary and only considers data in a given health zone and week, which corresponds to individual rows of our data frame.

::: {.write}
Add a `mutate()` to your pipeline, to create a `cases20` column that contains `1` if a given health zone has 20 cases or more in that week, and `0` otherwise.


The top of the data frame created by the pipe thus far looks like this:

```{r}
#| echo: false
#| eval: true

data_surv_weeks |>
  
  # Select the 4 health zones
  filter(health_zone %in% c("dilolo", "kowe" ,"kampemba", "lwamba")) |>
  
  # Order by provinceince, health zone and week
  arrange(province, health_zone, week) %>%
  
  # Binary indicator for 20 cases (weekly indicator by health zone)
  mutate(
    cases20 = case_when(
      totalcases >= 20 ~ 1, 
      .default = 0))  %>% head(10)
```
:::

#### Cumulative indicator

The second indicator you want to calculate is whether a health zone has more than 35 *cumulated suspected cases within three weeks*. This is a bit more complicated than the previous case: within heath zone you need to calculate the sum of cases by groups of three weeks, but the groups are not fixed: they are *rolling* across time. We are getting in the teritory of *moving* averages/sums/etc.

##### Cumulative sum

We are going to use the `rollapply()` function from the `{zoo}` package, as it is versatile and powerful. As its name suggests, the `rollapply()` function applies a function in a rolling way to a vector or a column of a data frame.

Since we are constrained in time, we are going to provinceide the code of the `rollapply()` function to calculate the cumulative sum over three weeks, but check out the details in the Going further section at the end when you have time. 

This is how to do it for *one health zone*:

```{r}
#| eval: true
example_df = data.frame(
  province    = "Haut Katanga",
  health_zone = "Dilolo",
  week        = 1:10,
  totalcases  = rep(1, times = 10))

example_df %>% 
  mutate(cumcas = rollapply(
    data  = totalcases,
    width = 3,          # Width of the window  
    FUN   = sum,        # Function to apply, here the sum   
    align = "right",    # Windows are aligned to the right
    partial = TRUE,     # Allows calcul to be made even if window is less than three
    
    na.rm = TRUE        # Extra unamed argument to be passed to the sum function
  )
  )
```

##### By health zone

Now, we want to do this cumulative sum *by health zone*. This is not that complicated: we are going to sort our data frame properly by health zone and week, and use the `.by` argument to tell the `mutate()` function to perform the action *by health zone*.

::: {.callout-note}
You may remember from the [aggregation session](../sessions_core/05_summary_table.qmd#sec-stratify) how se *summarized* by groups using the `.by` argument in the `summarize()` function. This is exactly the same idea, except that instead of returning *one value by group* (as `summarize()` does), we want to return one value per row (as `mutate()` does).

As a little reminder of how `summarize()` + `.by` work, here is how we would calculate the total number of patients and deceased by provinceince over the whole dataset:

```{r}
#| eval: true
data_surv_weeks %>% 
  summarize(
    .by = province,  # Do things by provinceince
    cases_tot = sum(totalcases, na.rm = TRUE),
    dead_tot = sum(totaldeaths, na.rm = TRUE)
  )
```

:::

::: {.write}
1. Add a step to your previous pipeline to sort the data frame by provinceince, health zone and week with the `arrange()` function.

2. Then add the following code to calculate the cumulative sum:

```{r}
mutate(
  .by = c(province, health_zone),
  cumcas = rollapply(
    data  = totalcases,
    width = 3,          # Width of the window  
    FUN   = sum,        # Function to apply, here the sum   
    align = "right",    # Windows are aligned to the right
    partial = TRUE,     # Allows calcul to be made even if window is less than three
    na.rm = TRUE        # Extra unamed argument to be passed to the sum function
  )
)
```

3. Add a new step to calculate a binary indicator, `cumcases35` that is `1` if the cumulative sum of cases for that week is equal or above 35 and `0` if not.

4. Add a new column `alert`, that is `1` if either the `cases20` indicator or the `cumcases35` indicator is `1` and `0` otherwise. You can use the `|` operator, which is R logical OR.

5. When the pipe is working, assign the result to a `data_alert` data frame.

`data_alert` should look like this:

```{r}
#| echo: false
#| eval: true
data_alert %>% head(10)
```

:::

### Health zones in alert (Q4)

After all this work we can *finally* investigate which health zones are in alert in the last week of our dataset (the *now* of the case study, week 20)!

::: {.write}
Filter your data frame to only keep the 20th week. Which health zones are in alert?


Create a vector `hz_alert` that contains the name of the health zones in alert, so that we can use it to filter data from these health zones later.
:::


## Draw the epicurve (Question 5)

Let us draw the epicurves of health zones currently in alert (in alert during week 20).

We have drawn very similar curves in the [epicurve session](../sessions_core/06_epicurves.qmd). Here again we will use the `ggplot()` function with the `geom_col()` geom to create a barplot showing the distribution of cases. Since we already have the number of cases per week we do not need to *count* it ourselved like we did in the past.


::: {.write}
Draw an epicurve for one of the health zone in alert. 

The graph should look like this (but maybe for another health zone):

```{r}
#| echo: false
#| eval: true
#| fig-width: 9
data_alert |>
  filter(health_zone == "kampemba") |>
  ggplot(aes(x = week, 
             y = totalcases)) + 
  geom_col(fill = "#2E4573") + 
  theme_bw(base_size = 15) + 
  labs(x = "Week",
       y = "N cases",
       title = "Kampemba health zone (in alert)")
```

:::

The `facet_wrap()` function allows us to plot several subplots in the same graph (see the [faceting satellite](../sessions_extra/faceting.qmd) for more information on faceting):

```{r}
#| eval: true

data_alert |>
  filter(health_zone %in% hz_alert) |>
  ggplot(aes(x = week, 
             y = totalcases)) + 
  geom_col(fill = "#2E4573") + 
  theme_bw(base_size = 15) + 
  labs(x = "Week",
       y = "N cases",
       title = "Health zones in alert") +
  facet_wrap(vars(health_zone))   # One graph by health_zone

```


## Key indicators (Question 6)

Let's gather more data on both alerts to help you decide which one to investigate.

::: {.callout-tip}
This session builds on summarizing skills seen in [the summary session](../sessions_core/05_summary_table.qmd). Do not hesitate to check it or your code if you forgot something.
:::


### Week of the first alert

::: {.write}
Use the `summarize()` function to display the first week the alert was raised for each health zone in alert. Which health zone started first?
:::


### Surveillance data indicators

Let us go back to the full surveillance dataset that contains more columns of interest.

::: {.write}
1. Add a column `cunder_5` to `data_surv` that contains the the number of cases less than five months.

2. Derive, for each health zone in alert, the following indicators (organized in a single table):

- The number of cases
- The number of deaths
- The number of less than five year olds
- The CFR in percentage
- The percentage of reported cases under five

The result should look like this: 

```{r}
#| echo: false
#| eval: true

data_surv |>
  filter(health_zone %in% hz_alert) |>
  mutate(
    cunder_5 = c011mois + c1259mois) %>%
  summarize(
    .by = health_zone,
    nb_cas       = sum(totalcases, na.rm = TRUE),
    nb_deces     = sum(totaldeaths, na.rm = TRUE),
    nb_under_5   = sum(cunder_5, na.rm = TRUE),
    cfr          = (nb_deces / nb_cas) * 100,
    prop_under_5 = (nb_under_5 / nb_cas) * 100
  )
```
:::


### Lab data indicators 

Now we are going to use the laboratory data to derive a couple more indicators.

::: {.write}
For each health zone in alert, derive the following indicators within one table:

- The number of patients tested for measles
- The number of positives for measles
- The percentage of positives for measles
- The number of patient tested for rubeole
- The number of positive for rubeole
- The percentage of positive for rubeole

```{r}
#| echo: false
#| eval: true

data_lab |>
  filter(health_zone %in% hz_alert) |>
  
  summarize(
    .by = health_zone,
    
    nb_roug_test  = sum(!is.na(igm_rougeole)),
    nb_roug_pos   = sum(igm_rougeole == "positif", na.rm = TRUE),
    prop_roug_pos = (nb_roug_pos / nb_roug_test) * 100,
    nb_rub_test   = sum(!is.na(igm_rubeole)),
    nb_rub_pos    = sum(igm_rubeole == "positif", na.rm = TRUE),
    prop_rub_pos  = (nb_rub_pos / nb_rub_test) * 100
  )
```
:::

::: {.callout-tip}
Check out the section on [summaries with conditions](../sessions_core/05_summary_table.qmd#sec-summary-cond) to remind you of the more advanced summaries.
:::


## Done!

Congratulation, you are done!

```{r}
#| echo: false
#| eval: true

downloadthis::download_link(
  link = 'https://github.com/epicentre-msf/repicentre/blob/main/solutions/extra/surveillance_module_solutions.R',
  button_label = 'Solution File',
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

## Going Further

### Exploring the `rollaply()` function 

If we want to do a cumulative sum of cases over three weeks, we want to apply the `sum()` function over windows of three weeks.

```{r}
#| eval: true
example_vect <- rep(1, time = 10)
example_vect

rollapply(
  data  = example_vect,
  width = 3,       # Width of the window  
  FUN   = sum,     # Function to apply, here the sum   
  align = "right"  # The final windows is aligned to the right
)
```

We inputed a vector of ten values and obtained a vector of lenght height, containing the sums. Obviously the function has a way of dealing with the extremities, and the size of the output is smaller than the size of the input. This would be a problem in a `mutate()` that creates new columns in a data frame, that need to be the same length as the existing columns.

You can control the behavior at the extremities:

- Fill with `NA` when there is not enough values to calculate a window of three
- Allow partial sums (some values represent less than three weeks)

The argument `fill = NA` pads the extremities with `NA` (on the left in our case, since we aligned right):

```{r}
#| eval: true
rollapply(
  data  = example_vect,
  width = 3,       # Width of the window  
  FUN   = sum,     # Function to apply, here the sum   
  align = "right", # Windows are aligned to the right
  fill = NA
)
```

It is a reasonnable way of dealing with incomplete windows. In our case however, we can do better: if there were 40 cases in week 1 it would be a cause for alert! We thus want the cumulative sum to be calculated from week one to be able to detect early alerts (keeping in mind that a *lack of alert* in the first two weeks may be a result of incomplete data). The `partial = TRUE` argument allows this:

```{r}
#| eval: true
rollapply(
  data  = example_vect,
  width = 3,       # Width of the window  
  FUN   = sum,     # Function to apply, here the sum   
  align = "right", # Windows are aligned to the right
  partial = TRUE)
```

This is close to what we need. 

A last point: you may remember that arithmetic operations in R return `NA` if some of the values are `NA` and we usually need to pass the argument `na.rm = TRUE` to the functions for them to ignore missing values.

If we had a slightly less complete vector we would have a problem:

```{r}
#| eval: true

example_vect_missing <- c(1, 1, 1, NA, 1, 1)

rollapply(
  data  = example_vect_missing,
  width = 3,       # Width of the window  
  FUN   = sum,     # Function to apply, here the sum   
  align = "right", # Windows are aligned to the right
  partial = TRUE   # Allows calcul to be made even if window is less than three
)
```

Fortunately we can pass the `na.rm = TRUE` argument to `rollapply()` so that it passes it to `sum()`.

```{r}
#| eval: true

rollapply(
  data  = example_vect_missing,
  width = 3,       # Width of the window  
  FUN   = sum,     # Function to apply, here the sum   
  align = "right", # Windows are aligned to the right
  partial = TRUE,  # Allows calcul to be made even if window is less than three
  na.rm = TRUE     # Extra unamed argument to be passed to the sum function
)
```

::: {.callout-tip}
Here we applied the `sum()` function to create a cumulative sum over 3 weeks. But you could, with minimal modifications, apply the `mean()` function to caclulate a moving average!
:::

### Pretifying percentages

The `percent` function from the `{scales}` packages can add percentage formatting to a value.

```{r}
scales::percent(0.8556)
```

It takes an `accuracy` arguments that controls the number of decimals:

```{r}
scales::percent(0.8556,
                accuracy = 0.1)
```

You can wrap it around the values that you calulate in the summary tables to change the proportions into nicely formatted percentages.